
In the following data set, we examine the effect of two diets on mice bodyweights. The variable `Diet` is a factor with two levels: "chow" and "hf." 
```{r echo = FALSE}
mice <- read.table(paste(my_dir,"/femaleMiceWeights.csv",sep=""),sep=",",header=TRUE)
```

```{r}
head(mice)
```

We fit a linear model in R and look at its design matrix $\mathbb{X}$.
```{r}
lm_mice <- lm(Bodyweight~Diet,data=mice)
model.matrix(lm_mice)
```
(a) Write down the sample linear model fitted in `lm_mice` using subscript format--this asks for the usual subscript format for linear models, not the double subscript format introduced to describe models with factors. Make sure to define appropriate notation.

```{r,child=paste(my_dir,"/sol-a.Rmd",sep=""),eval=SOL}
```

(b) In terms of the coefficients of this sample linear model, explain how to obtain estimates of the means of both treatment groups and the difference between these means.

```{r,child=paste(my_dir,"/sol-b.Rmd",sep=""),eval=SOL}
```
----------- 

