
To investigate the consequences of metal poisoning, 25 beakers of minnow larvae were exposed to varying levels of copper and zinc and the protein content was measured.
The data are as follows.
```{r, echo=F}
toxicity <- read.table(paste(my_dir, "/toxicity.txt",sep=""))
```

```{r lm_example, echo = F}
lm_toxicity <- lm(Protein~Copper+Zinc,data=toxicity)
round(coef(summary(lm_toxicity)), digits = 3)
```
The sample linear model is $\vect{y}=\mat{X}\vect{b}+\vect{e}$.
Here, $y_i$ is a measurement of total larva protein at the end of the experiment (in microgram, $\mu g$). $\mat{X}=[x_{ij}]$ is a $25\times 3$ matrix where $x_{i1}=1$, $x_{i2}$ is copper concentration (in parts per million, ppm) in beaker $i$ , and $x_{i3}$ is zinc concentration (in parts per million, ppm) in beaker $i$.

Suppose we're interested in predicting the protein in a new observation at 100ppm copper and 1000ppm zinc.

(a) Specify the values in a row matrix $\vect{x}^*$ such that $\vect{y}^*=\vect{x}^*\vect{b}$ gives a least squares prediction of the new observation. Calculate the predicted value.

```{r,child=paste(my_dir, "/sol_a.Rmd",sep=""),eval=SOL}
```

(b) Explain how to use the data vector $\vect{y}$, the design matrix $\mat{X}$, and your row vector $\vect{x}^*$ to construct a prediction interval that will cover the new measurement in approximately 95% of replications. Your answer should include formulas to construct this interval.

```{r,child=paste(my_dir, "/sol_b.Rmd",sep=""),eval=SOL}
```

(c) Calculate a 95% confidence interval for the relationship between zinc exposure and protein content in minnow larvae.

```{r,child=paste(my_dir, "/sol_c.Rmd",sep=""),eval=SOL}
```

(d)

```{r, echo = F, fig.height= 3, fig.width=3}
plot(lm_toxicity$fitted.values, toxicity$Protein,
     xlab = "fitted values", ylab = "Protein")
abline(0,1)
```

```{r, echo = F}
summary(toxicity)
```

Based on the graph above and the corresponding summary statistics, is this model a good fit for the data? Do you have any concerns about using this model for this prediction.

```{r,child=paste(my_dir, "/sol_d.Rmd",sep=""),eval=SOL}
```

-------


