\if \QLABELS1 Q2-1. \fi
To investigate the consequences of metal poisoning, 25 beakers of minnow larvae were exposed to varying levels of copper and zinc.
The data were

```{r lm_example}
toxicity <- read.table("toxicity.txt")
head(toxicity)
lm_toxicity <- lm(Protein~Copper+Zinc,data=toxicity)
round(coef(summary(lm_toxicity),3))
```
The sample linear model is $\vect{y}=\mat{X}\vect{b}+\vect{e}$.
Here, $y_i$ is a measurement of total larva protein at the end of the experiment (in microgram, $\mu g$). $\mat{X}=[x_{ij}]$ is a $25\times 3$ matrix where $x_{i1}=1$, $x_{i2}$ is copper concentration (in parts per million, ppm) in beaker $i$ , and $x_{i3}$ is zinc concentration (in parts per million, ppm) in beaker $i$.

Suppose we're interested in predicting the protein in a new observation at 100ppm copper and 1000ppm zinc.

(a) Specify the values in a row matrix $\vect{x}^*$ so that $\vect{y}^*=\vect{x}^*\vect{b}$ gives a least squares prediction of the new observation.


(b) Explain how to use the data vector $\vect{y}$, the design matrix $\mat{X}$, and your row vector $\vect{x}^*$ to construct a prediction interval that will cover the new measurement in approximately 95% of replications. Your answer should include formulas to construct this interval.

(c) Explain briefly some things you would look for to check whether your prediction interval is reasonable.

```{r,child="sol_q2-1.Rmd",eval=SOL}
```

-------


