\if \QLABELS1 Q2-3. \fi
We analyze the following data on video game sales in North America. This dataset records sales (in millions of dollars) for  580 games within three genres (shooter, sports and action) from two publishers (Electronic Arts and Activision) with years of release from 2006 to 2010 inclusive, on ten different platforms.
```{r}
vg <- read.table("vg_sales.txt") ; head(vg)
```
Consider the probability model $Y_{ijk} = \alpha + \beta_j + \gamma_k + \epsilon_{ijk}$ where $j=1,2,3$ specifies the genre (shooter, sports and action,  respectively), $k=1,2$ gives the publisher (Electronic Arts and Activision, respectively), and $i$ ranges over all the games in each $(j,k)$ category. In order to code these factors, we set $\beta_1=\gamma_1=0$.
As usual, $\epsilon_{ijk}$ gives an independent $N[0,\sigma]$ error for game $(i,j,k)$. Parameters in this probability model are estimated by least squares as follows:
```{r}
lm_vg1 <- lm(Sales ~ Publisher + Genre, data = vg)
summary(lm_vg1)
```
Note that the output of `summary(lm_vg1)` tells you that R is using $\vect{\beta}=(\alpha,\beta_2,\beta_3,\gamma_2)$ as the parameter vector.

(a) Write the first six lines of the design matrix $\mat{X}$ in the matrix version of the linear model $\vect{Y}=\mat{X}\vect{\beta}+\vect{\epsilon}$. Hint: the output from `head(vg)` tells you what the values of $j$ and $k$ are for each of the first six observations.


(b) Suppose we're interested in the predicting the North American Sales of a shooting game released by Activision. Specify a row matrix $\vect{x^*}$ such that $y^* = \vect{x^*b}$ gives the least square predictor of this quantity.

---------------

