\if \QLABELS1 Q4-1. \fi
We consider a dataset of measurements on crabs. The start of the dataset `crabs` is shown below. Here, `BD` refers to the body depth of the crabs. The species `sp` corresponds to the color of the crabs, which is a factor with two levels, Blue (`B`) and Orange (`O`). We want to study the difference of frontal lobe size (`FL`) of two species. 
```{r,echo=F}
library(MASS)
data(crabs)
```
```{r}
head(crabs)
```

Consider the probability model $Y_i = \mu_1x_{Bi} + \mu_2x_{Oi} + \epsilon_i$ for $i=1,...,200$. $Y_i$ is the frontal lobe size of crab $i$. $x_{Bi}$ is $1$ if crab $i$ is of species Blue and $0$ otherwise. Similarly, $x_{Oi}$ is $1$ if crab $i$ is of species Orange and $0$ otherwise. 
$\epsilon_i$ are i.i.d with mean $0$ and variance $\sigma^2$. This model can be fitted to the `crabs` dataset in R using the `lm()` function. The resulting summary is provided below.
```{r}
lm_crab <- lm(FL~sp-1, data=crabs)
summary(lm_crab)
```

(a) Interpret the meaning of $\mu_1$ and $\mu_2$ in the above probability model?

(b) Build a 95% confidence interval for $\mu_1$ using normal approximation

(c) Recall in homework we know that the full estimated covariance matrix of $\vect{\hat{\mu}} = (\hat{\mu}_1,\hat{\mu}_2)$ can be found by  
```{r}
V <- summary(lm_crab)$cov.unscaled * summary(lm_crab)$s^2
V
```
Use V and information provided in `summary(lm_crab)` to write down an expression that constructs a 95% confidence interval for $\mu_1-\mu_2$.

```{r,child="sol_q4-1.Rmd",eval=SOL}
```

------------------

