\if \QLABELS1 Q4-2. \fi
Consider the birth weight data set we have seen in lab.
For this question, we will look at columns `bwt` (birth weight), `lwt` (mother's weight), `age` (mother's age) and `race` (mother's race, 1 for white, 2 for black and 3 for other).
```{r}
library(MASS)
data(birthwt)
```

```{r}
head(birthwt)
```
We want to study the relationship between birthweight and race using an F test, while mother's weight and age are included as additional explanatory variables. Let the null hypothesis, $H_0$, be the probability model where birth weight is modeled to depend linearly on mother's weight and age. Let $H_a$ be the probability model where $H_0$ is extended to include race as a factor, as fitted in R by
```{r}
lm_bw <- lm(bwt ~ lwt + age +factor(race), data = birthwt)
```
The results from summary(lm_bw) and anova(lm_bw) are as follows
```{r}
summary(lm_bw)
```
```{r}
anova(lm_bw)
```

(a) Write out the null and alternative hypotheses of the F test by completely specifying the probability models.

```{r sol_q4-2a,child="sol_q4-2a.Rmd",eval=SOL}
```

(b) Interpret the results in `anova(lm_gpa)`. Specifically, 
read the sample test statistic from R output, give the distribution of the model-generated test statistic under $H_0$, and explain how the resulting p-value is calculated and interpreted. Your answer should givie relevant formulas, though you may define a residual sum of squares in words without explicitly saying how it is constructed.


```{r sol_q4-2b,child="sol_q4-2b.Rmd",eval=SOL}
```


---------

