---
title: "Quiz 2, STATS 401 F18"
author: "In lab on 11/16"
output:
  pdf_document:
  toc: no
header-includes:
   - \usepackage{ifthen}
csl: ecology.csl
---

\def\QLABELS{1}

```{r preliminaries,echo=F,cache=F}
ALL <- TRUE
#ALL <- FALSE
SOL <- FALSE
#SOL <- TRUE
library(mat2tex)
set.seed(5)
knitr::opts_chunk$set(cache = FALSE)
```


%% \newcommand\exam[1]{#1} %% exam paper formatting
\newcommand\exam[1]{} %% omit exam paper formatting


\newcommand\mat[1]{\mathbb{#1}}
\renewcommand\vec[1]{\boldsymbol{\mathrm{#1}}}
\newcommand\vect[1]{\vec{#1}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}
\newcommand\prob{\mathrm{P}}
\newcommand\E{\mathrm{E}}
\newcommand\SE{\mathrm{SE}}
\newcommand\normal{\mathrm{normal}}
\newcommand\MVN{\mathrm{MVN}}
\newcommand\RSS{\mathrm{RSS}}
\newcommand\pval{\mathrm{pval}}
\newcommand\sd{\mathrm{SD}}
\newcommand\sdSample{\mathrm{sd}}
\newcommand\var{\mathrm{Var}}
\newcommand\varSample{\mathrm{var}}
\newcommand\cov{\mathrm{Cov}}
\newcommand\covSample{\mathrm{cov}}



\exam{\vspace{1.5cm}

{\bf Name: \hspace{6cm} UMID: }

\vspace{1.5cm}

}

```{r,results="asis",echo=F,purl=F,eval=ALL}
cat("**PRELIMINARY VERSION. QUESTIONS NEED TO BE REWRITTEN AND/OR REARRANGED. This document produces different random quizzes each time the source code generating it is run. The actual quiz will be a realization generated by this random process, or something similar.**")
```

```{r,results="asis",echo=F,purl=F,eval=ALL}
cat("**This version lists all the questions currently in the quiz generator. The quiz will have one question drawn at random from each of the five categories. No new questions will be added after Wednesday 11/14. Small changes may be made.**")
```

**Instructions. You have a time allowance of 40 minutes, though the quiz may take you less time and you can leave lab once you are done.  The quiz is closed book, and you are not allowed access to any notes. Any electronic devices in your possession must be turned off and remain in a bag on the floor.**

### Formulas
  
The following formulas are provided. To use these formulas properly, you need to make appropriate definitions of the necessary quantities.

(1) $\quad\quad\mathbf{b}=\big(\mathbb{X}^\transpose \mathbb{X} \big)^{-1}\, \mathbb{X}^\transpose\mathbf{y}$

(2) $\quad\quad\var(X)=\E\big[ (X-\E[X])^2\big] = \E[X^2]-\big(\E[X]\big)^2$

(3) $\quad\quad$The probability density function of the standard normal distribution is $\frac{1}{\sqrt{2\pi}} e^{-x^2/2}$

(4) $\quad\quad$Syntax from `?pnorm`:
```
pnorm(q, mean = 0, sd = 1)
qnorm(p, mean = 0, sd = 1)
q: vector of quantiles.
p: vector of probabilities.
```

(5) $\quad\quad$If a random variable is normally distributed, the probability it falls within one standard deviation of the mean is 68%, within two standard deviations of the mean is 95%, and within three standard deviations of the mean is 99.7%.

-------------------------


\exam{\newpage}

### Q1. Normal approximations, mean and variance

```{r q1-all-list,include=F}
q1_files <- list.files(full.names=T,pattern="^q1") 
file.create("tmp1.Rmd")
file.append("tmp1.Rmd",q1_files)
```
```{r q1, child = sample(q1_files,size=1),eval=!ALL}
```
```{r q1-all, child = "tmp1.Rmd",eval=ALL}
```

\exam{\newpage}

### Q2. Covariance and the multivariate normal distribution
```{r q2-all-list,include=F}
q2_files <- list.files(full.names=T,pattern="^q2")
file.create("tmp2.Rmd")
file.append("tmp2.Rmd",q2_files)
```
```{r q2, child = sample(q2_files,size=1),eval=!ALL}
```
```{r q2-all, child = "tmp2.Rmd",eval=ALL}
```

\exam{\newpage}

### Q3. Prediction 

```{r q3-all-list,include=F}
q3_files <- list.files(full.names=T,pattern="^q3")
file.create("tmp3.Rmd")
file.append("tmp3.Rmd",q3_files)
```
```{r q3, child = sample(q3_files,size=1),eval=!ALL}
```
```{r q3-all, child = "tmp3.Rmd" ,eval=ALL}
```

\exam{\newpage}

### Q4. Linear models with factors

```{r q4-all-list,include=F}
q4_files <- list.files(full.names=T,pattern="^q4")
file.create("tmp4.Rmd")
file.append("tmp4.Rmd",q4_files)
```
```{r q4, child = sample(q4_files,size=1),eval=!ALL}
```
```{r q4-all, child="tmp4.Rmd",eval=ALL}
```

\exam{\vspace{10cm}}

------



License: This material is provided under an [MIT license](https://ionides.github.io/401w18/LICENSE)<br>




------
