The director of the CDC wants to assess how well rates of hospital-acquired infections (`Infection.risk`) can be predicted using properities of a hospital. She expects to use the average length of stay (`Length.of.stay`) in days, the average number of cultures for each patient without signs or symptoms of hospital-acquired infection, times 100 (`Culture`), the number of X-ray procedures divided by number of patients without signs or symptoms of pneumonia, times 100 (`X.ray`), and the number of beds a hospital has (`Beds`).

Let $\vect{x_1}$ be the length of stay, $\vect{x_2}$ be the culture count, $vect{x_3}$ be the number of X-rays, and $vect{x_4}$ is the number of beds. Consider the probability model $$Y_i = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + \beta_3x_{i3} + \beta_4x_{i4} + \epsilon_i$$ for $i = 1, \dots, 113$ where $\epsilon_i$ are iid normal($0, \sigma$).

She fits the linear model corresponding to this probability model in R:

```{r, echo=FALSE}
senic <- read.table("https://ionides.github.io/401w18/hw/hw09/senic.txt", header =T)
senic <- senic[,-1]
lm1 <- lm(Infection.risk ~ Length.of.stay + Culture + X.ray + Beds, data = senic)
round(summary(lm1)$coefficients[,1:2], 5)
```

(a) The CDC director asks you to determine if the size of the hospital (measured in the number of beds) affects the infection rate of the hospital. Write the null and alternative hypotheses we would use to answer this question.

```{r,child=paste(my_dir, "/sol_a.Rmd",sep=""),eval=SOL}
```

(b) What is the distribution of your test statistic from a?

```{r,child=paste(my_dir, "/sol_b.Rmd",sep=""),eval=SOL}
```

(c) Suppose we know that a local hopital has an average length of stay of 8 days, the average culture count is 14, the average number of X-rays is 90, and the number of beds is 40. Find the predicted value for this observation; you do not need to simplify.

```{r,child=paste(my_dir, "/sol_c.Rmd",sep=""),eval=SOL}
```

(d) Suppose we constructed a confidence interval for the expected infection rate for the hospital in part c. How would you check that your confidence interval is plausible?

```{r,child=paste(my_dir, "/sol_d.Rmd",sep=""),eval=SOL}
```

---------------

