---
title: "Stats 401 Lab 2"
author: "Sanjana Gupta"
date: "9/7/2018"
output:
  beamer_presentation:
    colortheme: dolphin
    incremental: no
  ioslides_presentation:
    incremental: no
fontsize: 18pt
---
# Office hours

- Dr. Ionides. Mon 10-11am, Wed 3-4pm (his office)

At 2165 USB, the Science Learning Center Annex

- Sanjana Gupta. Mon 4:30-5:30pm, Tu 11:30am-12:30pm
- Ed Wu. Tu 12:30-2:30pm
- Naomi Giertych. Thu 9-11 am

# Homework

- Out of 10 points
- $\textbf{0 points}$ if there is no statement of sources
- Provide the code if the question requests

# Swirl tutorial

We finished lessons 1/3/4 in HW1.

- Any techical difficulties encountered working with swirl?

- Any questions about materials introduced in the tutorial?



# Swirl tutorial

You are asked to complete lessons 5/6/7/9 for HW2. Lesson 9 can be a little bit harder. 

- We can go through parts of it together at the end of this lab (if we have time).

- You can always go to our office hours for help.
 
# Topics covered in today's Lab

- R functions  
    - R help: '?'
    - Apply function: 'apply()'  
- Summation notation

# R functions: help

- Access the documentation of functions by typing '?*name of function*'.
- Try '?mean', '?median'.
- What if you don't know the name of the inbuilt function?
  Try '??*name of concept*.
- Try '??columnnames', '??dimension'

# R functions: Apply

- Applies a given function to a vector or rows/ columns of a matrix.
- See documentation by typing '?apply'

Let's see an example: Find the average GPA and ACT scores of the students from the dataset used in the last lab.
```{r}
# Load the Dataset
gpa = read.table("CH01PR19.txt", header = T)

# Recall the dataset
#head(gpa)

# Find the mean of the columns
apply(gpa,2,mean)
```

# Summation

This is simply a compressed form of writing addition of many terms.

Given n constants $x_1, x_2, .., x_m, x_{m+1}, .., x_n$,
$$ \sum_{i=1}^n x_i = x_1 + x_2 + ... + x_{n-1} + x_n $$

In general,
$$ \sum_{i=m}^n x_i = x_m + x_{m+1} + ... + x_{n-1} + x_n $$

# Summation: Examples
- $\sum\limits_{i=1}^6 i = 1+2+3+4+5+6 = 21$

- $\sum\limits_{i=3}^5 i^2 = 3^2 + 4^2 + 5^2 = 9 + 16 + 25 = 50$

- $\sum\limits_{i=1}^9 1 = \underbrace{1+1+....+1}_{9 times} = 9$

- If $x_1$ = 11, $x_2$ = 22, $x_3$ = 21, $x_4$ = 12,
  $\sum\limits_{i=1}^4 x_i = x_1 + x_2 + x_3 + x_4 = 11 + 22 + 21+ 12 = 66$
  
Note: Comparing to basic summation formula on prev slide,  
$x_i = i$ in example 1, $x_i = i^2$ in example 2, $x_i = 1$ in example 3
  

# Summation: Basic Properties 
For any given (fixed) numbers n,m and constants c,d 

*Basic addition*

- $\sum\limits_{i=1}^n 1 = \underbrace{1+1+....+1}_{n times} = n\times 1 = n$  
- $\sum\limits_{i=1}^n d = \underbrace{d+d+....+d}_{n times} =n\times d$  
- $\sum\limits_{i=m}^n 1 = \underbrace{1+1+....+1}_{n-m+1 times} = (n-m+1)\times 1 = n-m+1$  
- $\sum\limits_{i=m}^n d = \underbrace{d+d+....+d}_{n-m+1 times} = (n-m+1)\times d = (n-m+1)d$


*Addition of summations*

- $\sum\limits_{i=1}^n x_i + \sum\limits_{i=1}^n y_i = \sum\limits_{i=1}^n (x_i + y_i)$
- $\sum\limits_{i=1}^4 (i + i^2) = \sum\limits_{i=1}^4 i + \sum\limits_{i=1}^4 i^2 = 10 + 30 = 40$


# Summation: Basic Properties (ctd)
 
*Scalar multiplication*

-  
$$
\begin{aligned}
c(\sum\limits_{i=1}^n x_i) &= c(x_1 + x_2 + ... + x_{n-1} + x_n)\\
                          &= cx_1 + cx_2 + ... + cx_{n-1} + cx_n\\
                          &= \sum_{i=1}^n c(x_i)
\end{aligned}
$$
                              
- $5\sum\limits_{i=1}^3 i = 5(1+2+3) =  5\times1 + 5\times2 + 5\times3 = \sum\limits_{i=1}^3 5i$

# Summation: Relating to linear model
Recall LM1, LM2 from ch1 notes:  
Suppose our data are $\{y_1,y_2,\dots,y_n\}$ and on each unit {i} we have {p} explanatory variables $\{x_{i1}, x_{i2},\dots,x_{ip}\}$. A linear model is  
for $i=1,2,\dots,n$
$$\tag{LM1} y_i = b_1 x_{i1} + b_2 x_{i2} + \dots + b_p x_{ip} + e_i$$ 
which is equivalent to
$$\tag{LM2} y_i = \sum_{j=1}^p x_{ij} b_j + e_i $$  

# In-lab Activity  
- Find the median GPA and ACT scores of the students from the dataset used in lab1 (CH01PR19.txt)  
- Express the mean of $x_1, x_2, \dots, x_n$ in summation notation.  
  *(Hint: Recall that $mean(x_1, x_2, \dots, x_n) = \frac{x_1 + x_2 + \dots + x_n}{n}$)*

# In-lab Activity Part 1
- Let us find the median GPA and ACT scores of the students.
```{r}
# Find the median scores 
apply(gpa,2,median)
```

# In-lab Activity Part 2
- Express mean in terms of summation
$$
\begin{aligned}
mean(x_1, x_2, \dots, x_n) &= \frac{x_1 + x_2 + \dots+ x_n}{n}\\
&= \sum\limits_{i=1}^n (x_i)/n\\
&= \frac{1}{n} \sum\limits_{i=1}^n x_i
\end{aligned}
$$

# Exit ticket
- Load the unemployment dataset from https://ionides.github.io/401f18/01/unemployment.csv  
- See the first few observations using the head() command  
- Find the average unemployment rate for each month  
  *(use apply and mean function)*  
- Recall the definition of standard deviation.  
  Find the inbuild function in R for standard deviation (using the help function '??') 
- Using this, find the standard deviation of the unemployment rate for each month  
