
\if \QLABELS1 Q3-3. \fi

(a) Which of the following is the matrix $\mat{A}$ generated by

```{r, eval = FALSE}
A <- t(matrix(c(rep(1,times=2),rep(3,times=2), 6, 4),ncol=3))
```

> (i) $\quad \mathbb{A} = \begin{bmatrix} 1 &  1 \\  3 &  3 \\ 6 & 4 \end{bmatrix}$

> (ii) $\quad \mathbb{A} = \begin{bmatrix} 1 &  3 & 6 \\  1 &  3 & 4 \end{bmatrix}$

> (iii) $\quad \mathbb{A} = \begin{bmatrix} 1 &  3 \\  1 &  6 \\ 1 & 3 \end{bmatrix}$

> (iv) $\quad \mathbb{A} = \begin{bmatrix} 1 &  1 & 3 \\  3 &  6 & 4 \end{bmatrix}$


```{r,child="sol_q3a-3.Rmd",eval=SOL}
```

(b) Which of the following successfully select the first five odd elements of the vector `x <- c(1,2,3,4,5,6,7,8,9,10,11)`?
(List all that apply. Do not list commands that will give an error)

> (i) $\quad$ `x[rep(c(TRUE,FALSE),each=5)]`

> (ii) $\quad$ `x[rep(c(TRUE,FALSE),times=5)]`

> (iii) $\quad$ `x[rep(c(TRUE,FALSE),length=9)]`

> (iv) $\quad$ `x[rep(c(TRUE,FALSE)][1:5]`

> (v) $\quad$ `x[rep(c("TRUE","FALSE"),5)]`

> (vi) $\quad$   None of the above

> (vii) $\quad$   All of the above

```{r,child="sol_q3b-3.Rmd",eval=SOL}
```

--------------------------


