
\if \QLABELS1 Q3-4. \fi

(a) 
Define the matrix `A` as:
```{r echo=FALSE}
A = matrix(c(0,1,1,3,3,2),nrow = 3)
A
```

What is the output of `apply(A,2,mean)`?

> (i). A vector of length 3 corresponding to the average of each row of `A`.

> (ii). A vector of length 2 corresponding to the average of each column of `A`.

> (iii). The mean of all the values in `A`. 

> (iv). The mean of the second column of `A`.

> (v). The mean of the second row of `A`.


```{r,child="sol_q3a-4.Rmd",eval=SOL}
```

(b)  For each of the lines of code below, say whether it will correctly make 50 draws from the $\normal(100,20)$ distribution. Among the correct answers, comment briefly on some strengths and weaknesses from the perspective of writing good R code. Which answer do you think is the best code, and why?

> (i) $\quad$ `rnorm(50,20,100)`

> (ii) $\quad$ `rnorm(100,20,50)`

> (iii) $\quad$ `rnorm(100,20,n=50)`

> (iv) $\quad$ `rnorm(mean=100,sd=20,n=50)`

> (v) $\quad$ `rnorm(n=50,mean=100,sd=20)`

> (vi) $\quad$ `replicate(rnorm(100,20),50)`

> (vii) $\quad$ `replicate(rnorm(n=1,mean=100,sd=20),n=50)`

> (viii) $\quad$ `rnorm(50)*20+100`

> (ix) $\quad$ `100+sqrt(20)*rnorm(50)
`

```{r,child="sol_q3b-4.Rmd",eval=SOL}
```

--------------------------



