
\if \QLABELS1 Q3-6. \fi

(a) 
Define the matrix `A` as:
```{r echo=FALSE}
A = matrix(c(1,2,3,6,1,2,4,1,1,3,3,6),nrow = 3)
A
```

What is the output of `apply(A[,-1],1,sd)`?

> (i). A vector of length 4 corresponding to the standard deviation of each column of `A`, excluding the first column.

> (ii). A vector of length 3 corresponding to the standard devation of each row of `A`, excluding the first column.

> (iii). The standard devation of all the values in `A`. 

> (iv). The standard devation of the first row of `A`.

> (v). An error since `A` doesn't have a -1 column.


```{r,child="sol_q3a-6.Rmd",eval=SOL}
```

(b)  Which of the following lines of code successfully contructs the matrix for part (a)? Comment on the strengths and weaknesses of the correct answers.

> (i). cbind(c(1,2,3), c(6,1,2), c(4,1,1), c(3,3,6), c(4,1,2))

> (ii). matrix(cbind(c(1,2,3), c(6,1,2), c(4,1,1), c(3,3,6), c(4,1,2)))

> (iii). t(matrix(c(1,6,4,3,4,2,1,1,3,1,3,2,1,6,2),nrow = 3))

> (iv). matrix(c(1,2,3,6,1,2,4,1,1,3,3,6,4),nrow = 3)


```{r,child="sol_q3b-6.Rmd",eval=SOL}
```

--------------------------



