\if \QLABELS1 Q4-2. \fi

Suppose we are interested in studying how well weight can be predicted by height. We collect the height and weight of 30 individuals and record them in the R vectors `height` and `weight`.

```{r, echo = F}
set.seed(401)
height = round(runif(100,58,76))
weight = height*1.5 + 40 + rnorm(100,0,10)
```

We fit a linear model and plot the data and the resulting fitted line using the following R code:
```{r,out.width="100mm",fig.width=6,fig.height=4}
weight_lm = lm(weight ~ height)
plot(height,weight)
abline(coef(weight_lm))
```

(a) Write out the linear model in subscript notation including the following coefficients from `weight_lm`. Make sure to define appropriate notation.

```{r, echo = F}
round(coef(weight_lm),3)
```

(b) Use matrix notation to explain how these coefficients were calculated.

(c) The height for observation $i = 10$ is 66. Write out the formula for the fitted value for this observation. You do not need to simplify your calculation.

(d) Use matrix notation to write out an expression for the fitted values of the model. Make sure to define appropriate notation.


```{r,child="sol_q4-2.Rmd",eval=SOL}
```


--------------

