
\if \QLABELS1 Q4-3. \fi

```{r, echo=F}
library(faraway)
data("sat")
sat <- sat[,c("expend","total")]
colnames(sat) <- c("expenditure","SAT")
```

The government wants to understand the relationship between expenditures on public education and test results. The dataset `SAT` contains the per-pupil annual expenditure (in thousands of dollars) and the average SAT score for each of the 50 states in 1994-95. 
The plot below shows a line fitted to a scatterplot of the points in the dataset.
```{r,out.width="100mm",fig.width=6,fig.height=4}
sat_lm <- lm(SAT~expenditure,data=sat)
plot(SAT~expenditure,data=sat)
abline(coef(sat_lm))
```

(a) Write out the regression model in subscript form (including an intercept term). Use letters, rather than actual numbers, for the model coefficients: you don't have the actual numbers at this point.
Make sure to define any notation you introduce.


```{r,child="sol_q4a-3.Rmd",eval=SOL}
```

\exam{\newpage}

The table below shows the head of the dataset (the first 6 rows).
```{r,out.width="100mm",fig.width=6,fig.height=4}
head(sat)
```

(b) Write out the corresponding design matrix. You only need to put in actual numbers for the first 5 rows, use ... after that and specify the dimension of the matrix.

```{r,child="sol_q4b-3.Rmd",eval=SOL}
```

\exam{\vspace{5cm}}

(c) Explain how the model coefficients are evaluated. Name the method and give the appropriate formula.

```{r,child="sol_q4c-3.Rmd",eval=SOL}
```

\exam{\vspace{4cm}}

(d) Describe in one line what trend you observe from the plot (what would you interpret from this data). Is this what you would've expected? What could be a possible justification for the trend being observed?

```{r,child="sol_q4d-3.Rmd",eval=SOL}
```

\exam{\vspace{4cm}}

----------------------------


