
\if \QLABELS1 Q4-3. \fi

```{r, echo=F}
library(faraway)
data("sat")
sat <- sat[,c("expend","total")]
colnames(sat) <- c("expenditure","SAT")
```

The government wants to study the relationship between expenditures on public education and test results in 1994-95. The following dataset contains the expenditure and the average SAT score for each of the 50 states. 

The plot below shows a line fitted to a scatterplot of the points in the dataset.
```{r}
sat_lm <- lm(SAT~expenditure,data=sat)
plot(SAT~expenditure,data=sat)
abline(coef(sat_lm))
```

(a) Write out the regression model in subscript form (including an intercept term).


The table below shows the head of the dataset (the first 6 entries).
```{r,out.width="100mm",fig.width=6,fig.height=4}
head(sat)
```

(b)Write out the corresponding design matrix. You only need to put in actual numbers for the first 5 rows, use ... after that and specify the dimension of the matrix.

(c) Explain how the model coefficients are evaluated. Name the method and give the appropriate formulae.

(d) Describe in one line what trend you observe from the plot (what would you interpret from this data). Is this what you would've expected? What could be a possible justification for the trend being observed?

----------------------------


