
\if \QLABELS1 Q4-4. \fi

A statistician employed by an avocado producer is asked to assess the relationship between avocado prices and sales volumn for small Hass avocados. She obtains data from \url{https://www.kaggle.com/neuromusic/avocado-prices} and reads them into R as a dataframe. She keeps only the 2016 data for organic avocados sold in the Detroit area and plots the average price in dollars ('AveragePrice') against the number of small Hass avocados sold ('X4046'). This results in a dataset with 52 observations.

```{r}

avocado <- read.csv("avocado.csv")
avocado_2016 <- subset(avocado, year == 2016 & type == 'organic' & region == 'Detroit')

```

She fits a linear model and plots the data and the resulting fitted line using the following R code:
```{r,out.width="100mm",fig.width=6,fig.height=4}

price_lm <- lm(X4046 ~ AveragePrice, data = avocado_2016)
plot(avocado_2016$AveragePrice, avocado_2016$X4046,
     xlab = 'Average Price', ylab = 'Number Sold')
abline(coef(price_lm))

```

(a) Write out the fitted linear model using subscript notation, including the following coefficients from `price_lm`. 
This means you are asked to use actual numbers, rather than letters, for the model coefficients.
Make sure to define any notation you introduce.

```{r echo = T}
round(coef(price_lm),3)
```

```{r,child="sol_q4a-6.Rmd",eval=SOL}
```

\exam{\newpage}

(b) Use matrix notation to explain how these coefficients were calculated.

```{r,child="sol_q4b-6.Rmd",eval=SOL}
```

\exam{\vspace{5cm}}

(c) Use matrix notation to write out an expression for the residual values of the model. Make sure to define appropriate notation.

```{r,child="sol_q4c-6.Rmd",eval=SOL}
```

\exam{\vspace{5cm}}

(d) From the scatter plot above, the statistician notices a potential outlier. This potential outlier corresponds to the the second week in January 2016. This week organic small Hass avocados sold for an average of $1.48 with a total of 4,763 sold. Write a numeric expression for the residual of this observation---you are not expected to evaluate it.

```{r,child="sol_q4d-6.Rmd",eval=SOL}
```

\exam{\vspace{5cm}}

--------------------------------


