\if \QLABELS1 Q5-1. \fi

The figure below shows the probability density function of a normal random variable $X$.  

```{r,echo=F,out.width="100mm",fig.width=6,fig.height=4}
pnorm_plot <- function(q_lo,q_hi,mean,sd,...) {
  coord.x <- seq(from=mean-3*sd,to=mean+3*sd,by=sd/100)
  plot(coord.x,dnorm(coord.x,mean,sd),type="l",...)
  shaded.coord.x <- coord.x[coord.x <= q_hi & coord.x >= q_lo]
  polygon.x <- c(min(shaded.coord.x),shaded.coord.x,max(shaded.coord.x))
  polygon.y <- c(0,dnorm(shaded.coord.x,mean,sd),0)  
  polygon(polygon.x,polygon.y,col='skyblue')
}
pnorm_plot(mean=50,sd=10,q_lo=70,q_hi=100,main="",xlab="",ylab="probability density")
```

(a) By looking at the probability density function, estimate the mean and standard deviation of $X$. Use these estimates for the subsequent parts of this question.

```{r,child="sol_q5a-1.Rmd",eval=SOL}
```

(b) Write a probability statement about the random variable $X$ that corresponding to the shaded area. 

```{r,child="sol_q5b-1.Rmd",eval=SOL}
```

(c) Write an integral corresponding to this shaded area. 

```{r,child="sol_q5c-1.Rmd",eval=SOL}
```

(d) Write R code to evaluate this integral numerically.

```{r,child="sol_q5d-1.Rmd",eval=SOL}
```

-------------
