
\if \QLABELS1 Q5-5. \fi

The figure below shows the probability density function of a normal random variable $X$.  

```{r,echo=F,out.width="100mm",fig.width=6,fig.height=4}
mean5.3 <- 50
sd5.3 <- 5
pnorm_plot_exterior <- function(q_lo,q_hi,mean,sd,...) {
  coord.x <- seq(from=mean-3*sd,to=mean+3*sd,by=sd/100)
  plot(coord.x,dnorm(coord.x,mean,sd),type="l",...)
  shaded.coord.x.lo <- coord.x[coord.x <= q_lo]
  shaded.coord.x.hi <- coord.x[coord.x >= q_hi]
  polygon.x.lo <- c(min(shaded.coord.x.lo),shaded.coord.x.lo,max(shaded.coord.x.lo))
  polygon.x.hi <- c(min(shaded.coord.x.hi),shaded.coord.x.hi,max(shaded.coord.x.hi))
  polygon.y.lo <- c(0,dnorm(shaded.coord.x.lo,mean,sd),0)  
  polygon.y.hi <- c(0,dnorm(shaded.coord.x.hi,mean,sd),0)  
  polygon(polygon.x.lo,polygon.y.lo,col='skyblue')
  polygon(polygon.x.hi,polygon.y.hi,col='skyblue')
}
pnorm_plot_exterior(
  mean=mean5.3,
  sd=sd5.3,
  q_lo=mean5.3-3*sd5.3/2,
  q_hi=mean5.3+3*sd5.3/2,
  main="",xlab="",ylab="probability density")
```

(a) By looking at the probability density function, estimate the mean and standard deviation of $X$. Use these estimates for the subsequent parts of this question.

```{r,child="sol_q5a-6.Rmd",eval=SOL}
```

\exam{\vspace{2cm}}

(b) Write a probability statement about the random variable $X$ corresponding to the shaded area. 

```{r,child="sol_q5b-6.Rmd",eval=SOL}
```

\exam{\vspace{2cm}}

(c) Write an integral corresponding to this shaded area. 

```{r,child="sol_q5c-6.Rmd",eval=SOL}
```

\exam{\vspace{3cm}}

(d) Write R code to evaluate this integral numerically.

```{r,child="sol_q5d-6.Rmd",eval=SOL}
```

\exam{\vspace{2cm}}

---------------------


