---
title: "Quiz 1, STATS 401 W18"
author: "In lab on 10/5"
output:
  pdf_document:
  toc: no
header-includes:
   - \usepackage{ifthen}
csl: ecology.csl
---

\def\QLABELS{1}

```{r preliminaries,echo=F,cache=F}
ALL <- TRUE
#ALL <- FALSE
#SOL <- FALSE
SOL <- TRUE
library(mat2tex)
set.seed(2)
knitr::opts_chunk$set(cache = FALSE)
```

\ifthenelse{\isodd{1}}{
 \newcommand\exam[1]{#1} %% exam paper
 \newcommand\solution[1]{}
}{ 
 \newcommand\exam[1]{} %% solutions
 \newcommand\solution[1]{ {\bf Solution}. {#1}}
} 


\newcommand\mat[1]{\mathbb{#1}}
\renewcommand\vec[1]{\boldsymbol{\mathrm{#1}}}
\newcommand\vect[1]{\vec{#1}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}
\newcommand\prob{\mathrm{P}}
\newcommand\E{\mathrm{E}}
\newcommand\SE{\mathrm{SE}}
\newcommand\normal{\mathrm{normal}}
\newcommand\MVN{\mathrm{MVN}}
\newcommand\RSS{\mathrm{RSS}}
\newcommand\pval{\mathrm{pval}}
\newcommand\sd{\mathrm{SD}}
\newcommand\sdSample{\mathrm{sd}}
\newcommand\var{\mathrm{Var}}
\newcommand\varSample{\mathrm{var}}
\newcommand\cov{\mathrm{Cov}}
\newcommand\covSample{\mathrm{cov}}


**This document produces different random quizzes each time the source code generating it is run. The actual quiz will be a realization generated by this random process, or something similar.**

```{r,results="asis",echo=F,purl=F,eval=ALL}
cat("**This version lists all the questions currently in the quiz generator. The quiz will have one question drawn at random from each of the five categories. No new questions will be added after Wednesday 10/3. Small changes may be made.**")
```

**Instructions. You have a time allowance of 40 minutes, though the quiz may take you less time and you can leave lab once you are done.  The quiz is closed book, and you are not allowed access to any notes. Any electronic devices in your possession must be turned off and remain in a bag on the floor.**

### Formulas
  
The following formulas will be provided. To use these formulas properly, you need to make appropriate definitions of the necessary quantities.

(1) $\quad\quad\mathbf{b}=\big(\mathbb{X}^\transpose \mathbb{X} \big)^{-1}\, \mathbb{X}^\transpose\mathbf{y}$

(2) $\quad\quad\var(X)=\E\big[ (X-\E[X])^2\big] = \E[X^2]-\big(\E[X]\big)^2$

(3) $\quad\quad$The probability density function of the standard normal distribution is $\frac{1}{\sqrt{2\pi}} e^{-x^2/2}$

(4) $\quad\quad$Syntax from `?pnorm`:
```
pnorm(q, mean = 0, sd = 1)
qnorm(p, mean = 0, sd = 1)
q: vector of quantiles.
p: vector of probabilities.
```

---------------


### Q1. Matrix exercises

```{r q1-all-list,include=F}
q1_files <- list.files(full.names=T,pattern="^q1") 
file.create("tmp1.Rmd")
file.append("tmp1.Rmd",q1_files)
```
```{r q1, child = sample(q1_files,size=1),eval=!ALL}
```
```{r q1-all, child = "tmp1.Rmd",eval=ALL}
```



### Q2. Summation exercises
```{r q2-all-list,include=F}
q2_files <- list.files(full.names=T,pattern="^q2")
file.create("tmp2.Rmd")
file.append("tmp2.Rmd",q2_files)
```
```{r q2, child = sample(q2_files,size=1),eval=!ALL}
```
```{r q2-all, child = "tmp2.Rmd",eval=ALL}
```


### Q3. R exercises

```{r q3-all-list,include=F}
q3_files <- list.files(full.names=T,pattern="^q3")
file.create("tmp3.Rmd")
file.append("tmp3.Rmd",q3_files)
```
```{r q3, child = sample(q3_files,size=1),eval=!ALL}
```
```{r q3-all, child = "tmp3.Rmd" ,eval=ALL}
```

### Q4. Fitting a linear model by least squares

```{r q4-all-list,include=F}
q4_files <- list.files(full.names=T,pattern="^q4")
file.create("tmp4.Rmd")
file.append("tmp4.Rmd",q4_files)
```
```{r q4, child = sample(q4_files,size=1),eval=!ALL}
```
```{r q4-all, child="tmp4.Rmd",eval=ALL}
```

### Q5. Probability exercises

```{r q5-all-list,include=F}
q5_files <- list.files(full.names=T,pattern="^q5")
file.create("tmp5.Rmd")
file.append("tmp5.Rmd",q5_files)
```
```{r q5, child = sample(q5_files,size=1),eval=!ALL}
```
```{r q5-all, child="tmp5.Rmd",eval=ALL}
```


------



License: This material is provided under an [MIT license](https://ionides.github.io/401w18/LICENSE)<br>




------
