---
title: "Stats 401 Lab 6"
author: "Ed Wu"
date: "10/12/2018"
output:
  beamer_presentation:
  colortheme: dolphin
incremental: no
ioslides_presentation:
  incremental: no
fontsize: 18pt
---

# Lab Activity (Part 1)
1. If $\text{Cor}(W,Z) = 0.5$, what is the correlation of $\text{Cor}(2W,Z+1)$?

2. Let $(X,Y)$ take the values $(0,1),(1,1),(1,2)$, each with probability 1/3
    - What is the covariance of $X$ and $Y$?
    - We take a sample of size 5: $(0,1), (0,1), (1,2), (1,1), (1,2)$. What is sample covariance?
    
# Lab Activity (Part 1) Solutions
## Question 1 Part 1

$$
\begin{aligned}
Cor(2W,Z+1) = \frac{Cov(2W, Z + 1)}{\sqrt{var(2W)var(Z+1)}} \\ \\
Cov(2W, Z + 1) = E[2W - E(2W)]E[(Z+1) - E(Z+1)] \\
= E[2W - 2E(W)]E[Z+1 - (E(Z) + 1)] \\
= 2E[W - E(W)]E[Z - E(Z)] \\
= 2Cov(W, Z)
\end{aligned}
$$

# Lab Activity (Part 1) Solutions, cont.
## Question 1 Part 1 (cont)

$$
\begin{aligned}
Var(2W) = 4Var(W) \\
Var(Z + 1) = Var(Z) \\
\end{aligned}
$$

$$
\begin{aligned}
Cor(2W,Z+1) = \frac{2Cov(W, Z)}{\sqrt{4Var(W)Var(Z)}} \\ \\
= \frac{Cov(W, Z)}{\sqrt{Var(W)Var(Z)}} \\ \\
= Cor(W,Z) = 0.5
\end{aligned}
$$

# Lab Activity (Part 1) Solutions, cont.
## Question 1 Part 2

$$
\begin{aligned}
cov(x,y) = \frac{1}{n-1}\sum_{i=1}^n(x_i - \bar{x})(y_i - \bar{y}) \\
= \frac{1}{4-1}\sum_{i=1}^4(x_i - \bar{x})(y_i - \bar{y}) \\
= \frac{1}{4}\sum_{i=1}^4(x_i - \bar{x})(y_i - \bar{y}) \\
\end{aligned}
$$

$$
\begin{aligned}
\bar{x} = \frac{3}{5} \\ \\
\bar{y} = \frac{7}{5}
\end{aligned}
$$

# Lab Activity (Part 1) Solutions, cont.
## Question 1 Part 2

$$
\begin{aligned}
cov(x,y) = \frac{1}{4}\sum_{i=1}^4(x_i - \frac{3}{5})(y_i - \frac{7}{5}) \\
= \frac{1}{4}[(0 - \frac{3}{5})(1 - \frac{7}{5}) + (0 - \frac{3}{5})(1 - \frac{7}{5}) + \\
(1 - \frac{3}{5})(2 - \frac{7}{5}) + (1 - \frac{3}{5})(1 - \frac{7}{5}) + \\
(1 - \frac{3}{5})(2 - \frac{7}{5})] \\
= \frac{1}{4}[(- \frac{3}{5})(- \frac{2}{5}) + (- \frac{3}{5})(- \frac{2}{5}) + \\
(\frac{2}{5})(\frac{3}{5}) + (\frac{2}{5})(- \frac{2}{5}) + \\
(\frac{2}{5})(\frac{3}{5})]
\end{aligned}
$$

# Lab Activity (Part 1) Solutions, cont.
## Question 1 Part 2 cont.

$$
\begin{aligned}
= \frac{1}{4}[\frac{6}{25} + \frac{6}{25} + \frac{6}{25} -\frac{4}{25} + \frac{6}{25}] \\
= \frac{1}{4} \times \frac{20}{25} \\
= \frac{5}{25} = \frac{1}{5}
\end{aligned}
$$

# Lab Activity (Part 2)
The scatterplot below was generated from a bivariate normal distribution with mean vector $(0,0)$ 
```{r, warning = FALSE, echo = FALSE,out.width="45mm",fig.width=4,fig.height=4,fig.align = 'center'}
library(mvtnorm)
data_ii = rmvnorm(1000, mean = c(0,0), sigma = matrix(c(1,-0.75,-0.75,1),nrow = 2))
plot(data_ii[,1],data_ii[,2],xlab = "x",ylab = "y")
```

Which of the following is the variance-covariance matrix?

1. $\begin{bmatrix} 1 & 0 \\ 0 & 1\end{bmatrix}$; 2. $\begin{bmatrix} 1 & 0.25 \\ 0.25 & 1\end{bmatrix}$; **3. $\begin{bmatrix} 1 & -0.75 \\ -0.75 & 1\end{bmatrix}$**

# Lab Activity (Part 2)
The scatterplot below was generated from a bivariate normal distribution with mean vector $(0,0)$ 
```{r, warning = FALSE, echo = FALSE,out.width="45mm",fig.width=4,fig.height=4,fig.align = 'center'}
library(mvtnorm)
data_ii = rmvnorm(1000, mean = c(0,0), sigma = matrix(c(1,0.2,0.2,1),nrow = 2))
plot(data_ii[,1],data_ii[,2],xlab = "x",ylab = "y")
```

Which of the following is the variance-covariance matrix?

1. $\begin{bmatrix} 1 & -0.2 \\ -0.2 & 1\end{bmatrix}$; **2. $\begin{bmatrix} 1 & 0.2 \\ 0.2 & 1\end{bmatrix}$**; 3. $\begin{bmatrix} 1 & 0.7 \\ 0.7 & 1\end{bmatrix}$

# Lab Activity (Part 2)
The scatterplot below was generated from a bivariate normal distribution with mean vector $(0,0)$ 
```{r, warning = FALSE, echo = FALSE,out.width="45mm",fig.width=4,fig.height=4,fig.align = 'center'}
library(mvtnorm)
data_ii = rmvnorm(1000, mean = c(0,0), sigma = matrix(c(1,0,0,1),nrow = 2))
plot(data_ii[,1],data_ii[,2],xlab = "x",ylab = "y")
```

Which of the following is the variance-covariance matrix?

**1. $\begin{bmatrix} 1 & 0 \\ 0 & 1\end{bmatrix}$**; 2. $\begin{bmatrix} 1 & 0.25 \\ 0.25 & 1\end{bmatrix}$; 3. $\begin{bmatrix} 1 & -0.75 \\ -0.75 & 1\end{bmatrix}$

# Lab Ticket
1. Why is $\begin{bmatrix} 4 & 0 \\ 0.25 & 4\end{bmatrix}$ not a valid variance-covariance matrix?

2. Let $(X,Y)$ be bivariate normal with mean $(6,4)$ and variance-covariance matrix $\mathbb{V} = \begin{bmatrix} 4 & 0 \\ 0 & 9\end{bmatrix}$. 
    - What are the mean and standard deviation of $Y$?
    - What is the covariance of $X$ and $Y$?
    
# Lab Ticket Solutions

1. It is not a valid variance-covariance because it is not symmetric.

2. (a.) The mean of Y is 4 and the standard deviation of Y is 3.

2. (b.) The covariance of $X$ and $Y$ is 0; this does not mean that they are independent.
