
We have been recruited by a California university to explore the relationship between water salinity, water oxygen, and water temperature. We have been given 60 years of oceanographic data collected from the California Current by the California Cooperative Oceanic Fisheries Investigations. Below is a snapshot of the data. (Source: https://www.kaggle.com/sohier/calcofi)

- Depthm: Depth in meters
- T_degC: Water temperture in degrees Celsius
- Salnty: Water Salinity in g of salt per kg of water
- 02ml_L: $O_2$ mixing ratio in ml/L

```{r, echo = F, include = FALSE}
bottle_2016 <- read.csv(paste(my_dir, '/bottle_2016.csv', sep = ""), header = TRUE, row.names = 1)
bottle_2016 <- na.omit(bottle_2016)
lm1 <- lm(T_degC ~ Depthm + Salnty + O2ml_L, data = bottle_2016)
```

We fit a linear model to the data; the results are shown below.

```{r, echo = F}
round(summary(lm1)$coefficients[, c("Estimate", "Std. Error")], 3)
```

Suppose we observed a new outcome $\vect{x^*}$

(a) Suppose we wanted to calculate a 95% confidence interval for the expected value of the new outcome. Write the expression for this calculation and define all terms.

```{r,child=paste(my_dir, "/sol_a.Rmd",sep=""),eval=SOL}
```

(b) Suppose instead, we wanted to calculate a 95% prediction interval for the new outcome. Write the expression for this calculation and define all terms.

```{r,child=paste(my_dir, "/sol_b.Rmd",sep=""),eval=SOL}
```

(c) How would you check that your confidence and prediction intervals are plausible?

```{r,child=paste(my_dir, "/sol_c.Rmd",sep=""),eval=SOL}
```

(d) Calculate the 95% confidence interval for the relationship between oxygen levels and water temperature.

```{r,child=paste(my_dir, "/sol_d.Rmd",sep=""),eval=SOL}
```

------------

