
We analyze the following data on video game sales in North America. This dataset records sales (in millions of dollars) for  580 games within three genres (shooter, sports and action) from two publishers (Electronic Arts and Activision) with years of release from 2006 to 2010 inclusive, on ten different platforms.
```{r,echo=F}
vg <- read.table(paste(my_dir,"/vg_sales.txt",sep=""))
```

```{r}
head(vg)
```

Let $\textbf{y} = (y_1, \dots, y_{580})$ be the sales of the games. Let $x_{i,1} = 1$ if game $i$ is published by Activision and 0 otherwise. Similarly, let $x_{i,2} = 1$ if game $i$ is published by Electronic Arts and 0 otherwise. 

In R, we fit the sample linear model given by $y_i = m_1x_{i,1} + m_2x_{i,2} + e_i$ for $i = 1, \dots, 580$. 
```{r}
lm_vg2 <- lm(Sales ~ Publisher-1, data = vg)
summary(lm_vg2)
```
 
(a) What do the coefficients in the summary above measure?

```{r,child=paste(my_dir,"/sol-a.Rmd",sep=""),eval=SOL}
```

\exam{\newpage}

(b) What is the design matrix used to fit the model? Write out the first 6 rows.

```{r,child=paste(my_dir,"/sol-b.Rmd",sep=""),eval=SOL}
```

\exam{\vspace{5cm}}

(c) Suppose we wish to fit the model $y_i = b_0 + b_1x_{i,1} + e_i$ for $i = 1, \dots, 580$. What is the value of $b_1$?
```{r,child=paste(my_dir,"/sol-c.Rmd",sep=""),eval=SOL}
```

\exam{\vspace{5cm}}


-----------------------




