---
title: "HW03 Solutions"
author: "STATS401 team"
date: "9/17/2018"
output: html_document
---

\newcommand\mat[1]{\mathbb{#1}}
\renewcommand\vec[1]{\boldsymbol{\mathrm{#1}}}
\newcommand\xtranspose{{\mbox{\tiny \textsc{t}}}}
\newcommand\transpose{{\scriptscriptstyle \mathrm{T}}}


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q2
### (a)

```{r, Q2(a1)}
# (a)
## download unemployment data file
download.file(destfile="unemployment.csv",
  url="https://ionides.github.io/401f18/01/unemployment.csv")

## read-in unemployment data into R
data <- read.table(file="unemployment.csv",sep=",",header=TRUE)
```

To define y, we look at a snipet of the dataset to see how to obtain 'y'.
```{r, echo=FALSE, Q2(a2)}
# (a) ctd
## look at dataset
head(data)
```

As we can see, to define 'y', we need to take the average of all columns except the first one. This can be done in one of the following two ways:
```{r, Q2(a3)}
# (a) ctd
## Defining y
y <- apply(data[,-1],1,mean)
  #or
y <- apply(data[,2:ncol(data)],1,mean)
```

### (b)
Recall from class notes (Notes03,LM3 and Notes03, slide15) that the design matrix for such a model will be $\mat{X}=[\mathbf{x_1} \mathbf{x_2]}$ where $\mathbf{x_1}$ is the vector of years (that is, its $i^{th}$ entry corresponds to the $i^{th}$ year in the dataset) and $\mathbf{x_2}$ is a vector of 1s (of length equal to the number of observations in the dataset). Let us now construct this matrix
```{r, Q2(b)}
#(b)
## Extract vector of years
x1 <- data$Year

## Define a vector containing 1s (of appropriate length)
x2 <- rep(1,nrow(data))

## Define X = [x1 x2]
X <- cbind(x1,x2)
```

### (c)
Recall formula for to obtain **b** from class notes (Notes03, LM4)
```{r, Q2(c)}
# (c)
b <- solve(t(X)%*%X)%*%t(X)%*%y
rownames(b) <- c("b1","b2") #renaming the elements of b, this is not essential, being done here so that the notation                              matches that of the hw
b
```

### (d)
Here, we have to rename the first column ($\mathbf{x_1}$) as $\mathbf{x}$.
```{r, Q2(d)}
# (d)
## Define x
x <- x1

## Define vector of fitted values
fitted_values <- c(X%*%b)

## Plot y ~ x
plot(x,y, main="Plot of average unemployment per year", ylab="average unemployment", xlab= "year")
lines(x,fitted_values,lty="solid",col="blue")
```

Clearly, there seems to be an increasing trend, that is, the average annual unemployment rate appears to be increasing every year.

### (e)
```{r, Q2(e)}
# (e)
## Fitting the linear model
lm1 <- lm(y~x1, data = data) #remember x is labeled as x1 in the design matrix

## Compare first 5-values
lm1$fitted.values[1:5]
fitted_values[1:5]

## Check if both vectors are equal by taking their difference
table(round(lm1$fitted.values - fitted_values,4))
```

## Q3
If $\mat{A},\mat{B}$ are two invertible square matrices, we have the property that $\mat{(AB)}^{-1} = \mat{B}^{-1}\mat{A}^{-1}$.  
However, this formula runs into difficulties if $\mat{A}$ or $\mat{B}$ is not invertible.
This is the problem in the argument.
In fact, $\mat{X}$ is usually not invertible. One way to see this is that $\mat{X}$ being invertible implies there is an exact solution to the system of equations $\mat{X}\vec{b}=\vec{y}$ given by $\vec{b}=\mat{X}^{-1}\vec{y}$. 
You may know that to determine $p$ variables in a system of linear equations, we need $p$ equations. Fewer than $p$ equations and we won't have enough information to solve for $p$ variables. 
More than $p$ equations and likely there won't be a solution.
Thus, only a square matrix is invertible. 
For us, $n$ is the number of data points and $p$ is the number of coefficients in the linear model so usually $p$ is much less than $n$. 
Thus, $\mat{X}$ is not square and cannot be invertible.

If $\mat{X}$ were invertible then (a) would be true. The following steps (b), (c) and (d) are correct matrix calculations and so this product would in fact equal identity in that case.

## Q4
Since $\mat{A}$ is a square matrix whose determinant is $6 \neq 0$, its inverse exists. 
Now, we know that $\mat{A}\mat{A}^{-1} = \mat{I}$ where $mat{I}$ is a diagonal matrix as well. Also, from above we know that $\mat{A}\mat{A}^{-1}$ is the same as multiplying the first columns of $\mat{A}^{-1}$ by 1,2, and 3 respectively. Hence, the natural guess would be that \[
\mat{A}^{-1} = \left[
\begin{array}{ccc}
1 & 0 & 0 \\
0 & 1/2 & 0 \\
0 & 0 & 1/3 
\end{array}
\right].
\]
and indeed this is correct! Let's verify using R.
```{r}
# Q4
## Define A, inv_A 
A <- matrix(c(1,0,0,0,2,0,0,0,3),3)
inv_A <- matrix(c(1,0,0,0,1/2,0,0,0,1/3),3)

## Verify 
A%*%inv_A
```

