Switzerland, in 1888, was entering a period known as the demographic transition; i.e., its fertility was beginning to fall from the high level typical of underdeveloped countries. This Swiss government has commissioned us to determining the factors most contributing to this decline.

We collect the following variables for each of the 47 French-speaking provinces around 1988: 

* Fertility: common standardized fertility measure
* Agriculture: % of males involved in agriculture as occupation
* Examination: % draftees receiving highest mark on army examination
* Education: % education beyond primary school for draftees.
* Catholic: % ‘catholic’ (as opposed to ‘protestant’).
* Infant.Mortality: live births who live less than 1 year. 

Let $\vect{x_1}$ be the agriculture rate, $\vect{x_2}$ be the examination rate, $\vect{x_3}$ be the education rate, $\vect{x_4}$ be the catholic rate, and $\vect{x_5}$ be the infant mortality rate. Consider the probability model 
\[
Y_i = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + \beta_3x_{i3} + \beta_4x_{i4}+ \beta_5x_{i5} + \epsilon_i
\]
for $i = 1, \dots, n$ with $n=47$ and $\epsilon_1,\dots,\epsilon_n\sim\iid\;\normal(0, \sigma)$.

We fit a the regression model corresponding to this probability model in R:


```{r,echo=F}
data(swiss)
lm1 <- lm(Fertility ~ ., data = swiss)
round(summary(lm1)$coefficients[,1:2], 3)
```


(a) The Swiss governmment is skeptical that the examination percentage affects the fertility rate. Write the null and alternative hypotheses we would use to answer this question.

```{r,child=paste(my_dir, "/sol_a.Rmd",sep=""),eval=SOL}
```

(b) (i) What is your test statistic for part (a)? (ii) What is the distribution of a model-generated test statistic under the null hypothesis? (iii) What is your conclusion for the hypothesis test in part (a)? No calculations are necessary.

```{r,child=paste(my_dir, "/sol_b.Rmd",sep=""),eval=SOL}
```

(c) A new province is conquered in 1889 and its statistics are added to our data. This new province had an agriculture rate of 70%, examination rate of 22%, and education rate of 10%, a catholic rate of 50%, and an infant mortality rate of 20%. Find a numeric expression for the predicted fertility rate of this new province.

```{r,child=paste(my_dir, "/sol_c.Rmd",sep=""),eval=SOL}
```

(d)

```{r, echo = F, fig.height=4, fig.width=4}
plot(lm1$fitted.values, swiss$Fertility,
     xlab = "Fitted Values", ylab = "Fertility")
abline(0,1)
```

```{r}
summary(swiss)
```

Based on the graph above and the corresponding summary statistics, is this model a good fit for the data? Do you have any concerns about using this model for this prediction.

```{r,child=paste(my_dir, "/sol_d.Rmd",sep=""),eval=SOL}
```

-------


