
We consider a subset of the National Education Longitudinal Study of 1988 which examined schoolchildren’s performance on a math test score in 8th grade. `ses` is the socioeconomic status of parents and `paredu` is the parents highest level of education achieved (less than high school, high school, college, BA, MA, PhD). The dataset called `nels88` starts as follows:
```{r,echo=F}
library(faraway)
data(nels88)
```
```{r lm_nels88_example}
head(nels88)
```
We fit a regression model to the data. The rounded co-efficients for the model are provided below:
```{r}
fit <- lm(math ~ ses + paredu, data = nels88)
round(summary(fit)$coef)
```

(a)  Describe a suitable probability model, in matrix form, to give a sample version of the linear model that has been fit above.

```{r,child=paste(my_dir,"/sol-a.Rmd",sep=""),eval=SOL}
```


(b) Find the predicted math score for a student whose family has an ses value of -0.5 and whose parents' highest education level is high school (`hs`).

```{r,child=paste(my_dir,"/sol-b.Rmd",sep=""),eval=SOL}
```


(c) How is the residual standard error calculated for this model? (Give a formula).

```{r,child=paste(my_dir,"/sol-c.Rmd",sep=""),eval=SOL}
```

-------


