We are interested in studying the relationship between the miles per gallon of a car and the number of cylinders its engine has. In the following data set, `mpg` corresponds to the miles per gallon of each car. The variable `cylinders` corresponds to the number of cylinders and takes the values "4 cyl", "6 cyl", or "8 cyl." The variable `horsepower` corresponds to the horse power of each car.

```{r, echo=F}
set.seed(2018)
mpg = read.table("auto-mpg")[,c(1,2,4)]
names(mpg) = c("mpg","cylinders","horsepower")
mpg = mpg[-which(mpg$cylinders == 3 | mpg$cylinders == 5),]
mpg = mpg[-which(is.na(mpg$mpg))]

mpg$cylinders = paste(mpg$cylinders,"cyl")
mpg$cylinders = as.factor(as.character(mpg$cylinders))
mpg = mpg[sample(1:nrow(mpg)),]
rownames(mpg) = 1:nrow(mpg)
```

```{r}
head(mpg)
```

Let $\textbf{x}_{1}$ be a dummy variable for 6 cylinder cars, $\textbf{x}_{2}$ be a dummy variable for 8 cylinder cars, and $\textbf{x}_{3}$ be horsepower. Consider the probability model $$Y_i = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + \beta_3x_{i3} + \epsilon_i$$ for $i = 1, \dots, 399$ where $\epsilon_i$ are iid normal($0, \sigma$). We fit the linear model corresponding to this probability model in R:

```{r}
lm_mpg = lm(mpg ~ cylinders + horsepower, data = mpg)
summary(lm_mpg)$coefficients[,1:2]
```

(a) What is the design matrix $\mathbb{X}$? Write out the first 6 rows.

```{r,child=paste(my_dir,"/sol-a.Rmd",sep=""),eval=SOL}
```

(b) Suppose we have a new car that has 6 cylinders and a horsepower of 110. What is the predicted miles per gallon? You do not need to simplify your calculation.

```{r,child=paste(my_dir,"/sol-b.Rmd",sep=""),eval=SOL}
```

(c) We want to know if 8 cylinder cars have lower miles per gallon on average than 4 cylinder cars (after controlling for horsepower). What are the null and alternative hypotheses we would use to answer this question?


```{r,child=paste(my_dir,"/sol-c.Rmd",sep=""),eval=SOL}
```


----------- 

