We consider a dataset of measurements on crabs. The start of the dataset `crabs` is shown below. The species `sp` corresponds to the color of the crabs, which is a factor with two levels, Blue (`B`) and Orange (`O`). We want to study the difference between the frontal lobe size (`FL`) of the two species. 
```{r,echo=F}
library(MASS)
data(crabs)
```
```{r}
head(crabs)
```

Consider the probability model $Y_i = \mu_1x_{Bi} + \mu_2x_{Oi} + \epsilon_i$ for $i=1,...,200$. $Y_i$ is the frontal lobe size of crab $i$. $x_{Bi}$ is $1$ if crab $i$ is of species Blue and $0$ otherwise. Similarly, $x_{Oi}$ is $1$ if crab $i$ is of species Orange and $0$ otherwise. 
$\epsilon_i$ are i.i.d with mean $0$ and variance $\sigma^2$. This model can be fit to the `crabs` dataset in R using the `lm()` function. The resulting summary is provided below.
```{r}
lm_crab <- lm(FL~sp-1, data=crabs)
summary(lm_crab)$coefficients[,1:2]
```

(a) Interpret the meaning of $\mu_1$ and $\mu_2$ in the above probability model

```{r,child=paste(my_dir,"/sol-a.Rmd",sep=""),eval=SOL}
```

\exam{\vspace{3cm}}

(b) Build a 95% confidence interval for $\mu_1$ using the normal approximation. You do not need to simplify your upper and lower bounds.

```{r,child=paste(my_dir,"/sol-b.Rmd",sep=""),eval=SOL}
```

\exam{\newpage}

(c) What is the design matrix used to fit the model above? Write out the first 6 rows. 

```{r,child=paste(my_dir,"/sol-c.Rmd",sep=""),eval=SOL}
```

\exam{\vspace{10cm}}


------------------

